% logtables.tex - CC BY-NC-SA - scruss.com - 2012-11-04

\documentclass[twoside,a4paper]{article}
\usepackage{longtable}
\usepackage{color}
\usepackage{colortbl}
\usepackage{graphicx}
\pagestyle{myheadings}
\usepackage[condensed,math]{iwona}
\usepackage[T1]{fontenc}
\usepackage{hyperref}

\hypersetup{
  pdftitle={Russell's Modern Logarithms},
  pdfauthor={Stewart C. Russell},
  pdfsubject={Logarithm and antilogarithm tables},
  pdfkeywords={logarithm} {antilogarithm} {multiplication} {division} {roots} {exponentiation}
}

\renewcommand{\tabcolsep}{3pt} % narrow table
\definecolor{bg}{RGB}{213,253,220} % alternate table row colour

\title{Modern Logarithms}
\author{Stewart C. Russell}

\begin{document}
\maketitle
\tableofcontents

\newpage
\thispagestyle{empty}
\section{Decimal Logarithms}
\newpage


% headings for log table
\makeatletter
\def\@oddhead{\thepage\qquad\textsc{Decimal Logarithms}\hfill \rightmark{} -- \leftmark}
\def\@evenhead{\rightmark{} -- \leftmark \hfill \textsc{Decimal Logarithms}\qquad\thepage}
\makeatother

\begin{longtable}[c]{|*{11}{c|}}
 & \textbf{0} & \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} & \textbf{6} & \textbf{7} & \textbf{8} & \textbf{9} \\
\hline \endhead
\hline \endfoot
%%%
\input{logbody}
\end{longtable}

\newpage
\thispagestyle{empty}
\section{Decimal Anti-Logarithms}
\newpage

% headings for antilog table
\makeatletter
\def\@oddhead{\thepage\qquad\textsc{Decimal Anti-Logarithms}\hfill \rightmark{} -- \leftmark}
\def\@evenhead{\rightmark{} -- \leftmark\hfill \textsc{Decimal Anti-Logarithms}\qquad\thepage}
\makeatother

\begin{longtable}[c]{|*{11}{c|}}
 & \textbf{0} & \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} & \textbf{6} & \textbf{7} & \textbf{8} & \textbf{9} \\
\hline \endhead
\hline \endfoot
%%%
\input{antilogbody}
\end{longtable}

\newpage
\markboth{}{}
\makeatletter
\def\@oddhead{\thepage \hfill \rightmark{} \leftmark}
\def\@evenhead{\rightmark{} \leftmark\hfill \thepage}
\makeatother

\section{How to use this book}

Tables of logarithms (``log tables'') are a quick way of simplifying
multiplication and division tasks when no calculator is available. This is due to the simple identities:

\begin{itemize}
\item $\log{(a \times b)} = \log{a} + \log{b}$
\item $\log{(a \div b)} = \log{a} - \log{b}$
\item $\log{a^b} = \log{(a)} \times b$
\item $\log{\sqrt[b]{a}} = \log{(a)} \div b$ 
\end{itemize}

To keep this book a reasonable size, logarithms are given for numbers in the range 1.000 -- 9.999. In order to work with larger or smaller numbers, it is necessary to convert them to scientific notation, $a \times 10^b$, where $1.000 \leq a < 9.999$. For example, 21.8 becomes  $2.18 \times 10^1$ and 0.00218 becomes $2.18 \times 10^{-3}$. It's important to keep track of the powers of ten, as in multiplication and division they control the position of the decimal point, and in roots they become involved in the table lookup process.

To look up a logarithm, the number is found in the left hand index row to two decimal places, and then the correct column is selected for the third decimal place. For example, to find the logarithm of 3.657:

\begin{longtable}[c]{|*{11}{c|}}
 & \textbf{0} & \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} & \textbf{6} & \textcolor{red}{\textbf{7}} & \textbf{8} & \textbf{9} \\
\hline \endhead
\hline \endfoot
%%%
 \textbf{3.64} & \multicolumn{10}{c|}{...} \\
\rowcolor{bg} \textcolor{red}{\textbf{3.65}} & .56229 & .56241 & .56253 & .56265 & .56277 & .56289 & .56301 & \textcolor{red}{\textit{.56312}} & .56324 & .56336 \\
 \textbf{3.66} & \multicolumn{10}{c|}{...} \\
\end{longtable}

Thus $\log{3.657} = 0.56312$. This is more correctly written $\log_{10}{3.657} = 0.56312$, but the decimal base will be omitted throughout.

To look up an anti-logarithm, a similar process is used. Antilogs are given to fewer decimal places that logarithms, again to keep the size of this book reasonable. To continue with our example, we will look up  $\log^{-1}{0.56312}$:


\begin{longtable}[c]{|*{11}{c|}}
 & \textbf{0} & \textcolor{red}{\textbf{1}} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} & \textbf{6} & \textbf{7} & \textbf{8} & \textbf{9} \\
\hline \endhead
\hline \endfoot
%%%
 \textbf{.562} & \multicolumn{10}{c|}{...} \\
\rowcolor{bg} \textcolor{red}{\textbf{.563}} & 3.6559 & \textcolor{red}{\textit{3.6568}} & 3.6576 & 3.6585 & 3.6593 & 3.6602 & 3.6610 & 3.6618 & 3.6627 & 3.6635 \\
 \textbf{.564} & \multicolumn{10}{c|}{...} \\
\end{longtable}

Note that we've had to round the lookup to 0.5631. It should surprise no-one that $\log^{-1}{0.5631} = 3.6568$, or our original example 3.657 when expressed to the same number of decimal places.

\subsection{Multiplication}

Example: $21.8 \times 36.57 = $?
\begin{enumerate}
\item Convert both numbers into scientific form: $2.18 \times 10^1 \times 3.657 \times 10^1$
\item Look up $\log{2.18}$. It's in the first column next to 2.18: \textbf{0.33846}
\item Look up $\log{3.657}$. It's in the eighth column next to 3.65: \textbf{0.56312}
\item Add the results: $0.33846 + 0.56312 = 0.90158$
\item Look up the anti-logarithm of 0.9016: \textbf{7.9726}
\item Multiply the result by the powers of ten left over: $7.9726 \times 10^{1+1} = 797.26$
\item Just by looking up logarithms and adding, $21.8 \times 36.57 \approx \textbf{797.26}$. Compare this with the calculated result, \textbf{797.226}.
\end{enumerate}

\subsection{Division}

Example: $23.27 \div 14.07 =$?
\begin{enumerate}
\item Convert both numbers into scientific form: $(2.327 \times 10^1) \div (1.407 \times 10^1)$
\item Look up $\log{2.327}$. It's in the eighth column next to 2.32: \textbf{0.36680}
\item Look up $\log{1.407}$. It's in the eighth column next to 1.40: \textbf{0.14829}
\item Subtract the results: $0.36680 - 0.14829 = 0.21851$
\item Look up the anti-logarithm of 0.2185: \textbf{1.6539}
\item Multiply the result by the powers of ten left over: $1.6539 \times 10^{1-1} = 1.6539$. (Since $n^0=1$.)
\item Just by looking up logarithms and subtracting, $23.27 \div 14.07 \approx \textbf{1.6539}$. Compare this with the calculated result, \textbf{1.654} to 4 significant figures.
\end{enumerate}


\subsection{Exponentiation}

Example: $8.232^3 = $?
\begin{enumerate}
\item Look up $\log{8.232}$. It's in the third column next to 8.23: \textbf{0.91551}
\item Multiply $0.91551 \times 3 = 2.74653$.
\item Split the result into integer (2) and fractional parts, and look up the anti-logarithm of the fraction, 0.7465: \textbf{5.5783}
\item Multiply the result by the integer part as powers of ten: $5.5783 \times 10^2 = 557.83$.
\item Just by looking up logarithms and multiplying, $8.232^3 \approx \textbf{557.83}$. Compare this with the calculated result, \textbf{557.8} to 4 significant figures.
\end{enumerate}

\subsection{Roots}

Example: $\sqrt[3]{28.317}= $?
\begin{enumerate}
\item Convert the number to scientific form: $2.8317 \times 10^1$, so the calculation becomes 
$\sqrt[3]{2.8317} \times \sqrt[3]{10}$
\item Look up $\log{2.832}$: \textbf{0.45209}
\item Divide $0.45209 \div 3 = 0.1507$.
\item We still have to deal with the $\sqrt[3]{10}$ term: $\log{10} = 1$, so $1 \div 3 = 0.3333$
\item Add the two results: $0.1507 + 0.3333 = 0.4840$
\item Look up the anti-logarithm of 0.4840: \textbf{3.0479}
\item So $\sqrt[3]{28.317} \approx 3.0479$. Compare this with the calculated result, \textbf{3.0480} to 5 significant figures.
\end{enumerate}
(If it is one's misfortune to have to work with ``imperial''
measurements, it will be quickly recognized that there are 28.317
litres [cubic decimetres] in a cubic foot, and 3.048 decimetres to the
foot.) 

\vfill

\end{document}  